---
title: "Nested Data - Homework"
author: "please insert your names here"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tasks

due till 2022-12-08

Use the data from the article [Determinants of healthcare worker turnover in intensive care units: A micro-macro multilevel analysis](https://doi.org/10.1371/journal.pone.0251779) by Daouda, Hocine & Temime (2021).

**Test the hypotheses:**

The stress level is affected by:

**H1**: L1 variable `support from supervisors`

**H2**: L2 variable `staff-to-patient ratio day`

Tip: you can reuse code from code that generated today's slides
# Preparations

## Libraries

Load required libraries

```{r load_libraries, message=FALSE}
# just run this chunk

library(tidyverse)
library(magrittr)
library(plotly)
library(kableExtra)
library(datapasta)
library(lme4)
library(lmeresampler)
```


## Data

1. get the files
  -  [Individual-level factors data](https://doi.org/10.1371/journal.pone.0251779.s010)
  - [Intensive care unit-level factors data](https://doi.org/10.1371/journal.pone.0251779.s011)
2. add the folder `/journal.pone.0251779` into `/data` and copy both files into it


# Task

## Prep

Get data into shape.

```{r prepare_data}
# just run this chunk

df_individual <- read_csv2('../data/journal.pone.0251779/journal.pone.0251779.s010.csv')
df_unit <- read_csv2('../data/journal.pone.0251779/journal.pone.0251779.s011.csv')

df_ICU <- df_individual %>%
  left_join(df_unit) %>%
  mutate(
    CodeService = as_factor(CodeService),
    `type of ICU` = as_factor(`type of ICU`)
    )
```

Take a look at data to see how it is structured

```{r see_data}
# just run this chunk

df_ICU %>%
  # limit to top 20 rows
  slice_head(n=20)
```

## Optional

Explore the data (this is optional for the homework).

```{r}
# ADD YOUR CODE


```

## Test **H1** and **H2**.

Tip 1: you will need `lme4::lmer()` models
Tip 2: you can use the code from the lecture as a starting point.


```{r}
# ADD YOUR CODE

```


```{r}
# ADD YOUR CODE

```


```{r}
# ADD YOUR CODE

```

... add more code chunks as needed.


### Question

Please report your results.


### Answer

[YOUR ANSWER]



# At last

* check that you have
  * added your name at the top of the document as author
  * added all missing code segments
  * answered all questions
* knit to PDF or HTML
* upload to moodle