---
title: "Mean Differences - Homework"
author: "please insert your names here"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tasks

* TASK 1 (2-factors) 
  - Is there any Swiss region, where women/men are especially happy?
  - use the variables `H1` (happy/unhappy), `Nuts2` (region) and `DEMO1` (gender) from the [MOSAiCH](https://doi.org/10.48573/t659-e039) data set
  
* TASK 2 (mixed)
  - Can you reproduce the result regarding _psychological distress_ in [Prudenzi, et al. (2022)](https://doi.org/10.1371/journal.pone.0266357)? 
  - The data is available [here](https://doi.org/10.17605/OSF.IO/ZNCWA).
  - Would the result change with a robust approach?

# Preparations

## Libraries

Load required libraries

```{r load_libraries}
# just run this chunk

library(tidyverse)
library(WRS2)
```


## Data

1. Make sure that the [MOSAiCH](https://doi.org/10.48573/t659-e039) data-set is extracted at _[your individual folder path]/2022_SA_advanced_statistical_methods/data_.
2. Make sure that the [ACT for NHS staff data](https://doi.org/10.17605/OSF.IO/ZNCWA) is extracted at _[your individual folder path]/2022_SA_advanced_statistical_methods/data_.


# Task 1

## Prep

Get data into shape.

```{r prepare_mosaich_data}
# just run this chunk

df_mosaich <- haven::read_sav('../data/MOSAiCH_2021/swissubase_2033_1_0/2033_MOSAiCH2021_Data_E_v1.0.0.sav') %>%
  # use labels instead of values for Nuts2
  mutate(
    Nuts2 = haven::as_factor(Nuts2),
    DEMO1 = haven::as_factor(DEMO1)
    ) %>%
  # reduce to needed variables
  select(IDNO, Nuts2, DEMO1, H1) %>%
  # remove missings
  drop_na() %>%
  mutate(
    IDNO = as_factor(IDNO),
    # remove missing factor levels
    Nuts2 = forcats::fct_drop(Nuts2),
    DEMO1 = forcats::fct_drop(DEMO1),
    H1 = unclass(H1)
    )
```

Take a look at data to see how it is structured

```{r see_mosaich_data}
# just run this chunk

df_mosaich %>%
  # limit to top 20 rows
  slice_head(n=20)
```

## Optional

Explore the data (this is optional for the homework).

```{r}
# ADD YOUR CODE


```


## Parametric

We start by running a parametric 2-way Anova

```{r}
# just run this chunk
para_2way <- aov(H1 ~ Nuts2*DEMO1, data = df_mosaich)
para_2way
```

Let's check the parameter results

```{r}
# just run this chunk
para_2way %>% summary()
```

### Question

1. Please interpret the results.
2. Which parameter relates to the research question "Is there any Swiss region, where women/men are especially happy?" and how would you answer it based on the results?

### Answer

1. [YOUR ANSWER]
2. [YOUR ANSWER]


## Robust

Choose a robust version of the parametric test above in the [WRS2](https://cran.r-project.org/web/packages/WRS2/WRS2.pdf) and carry it out.

Start with the omnibus test.

```{r robust_2way_omnibus}
# YOUR CODE


```

Also conduct a robust post-hoc test.

```{r robust_2way_postHoc}
# YOUR CODE

```


_Note_: WRS2 does not print the labels of levels but just their number. To facilitate your interpretation, the code below lists the labels with the according number.

```{r levels_nuts2}
# just run this chunk
df_mosaich %>%
  select(Nuts2) %>%
  distinct() %>%
  mutate(number = as.numeric(as.factor(Nuts2))) %>%
  arrange(number)
```

```{r levels_demo1}
# just run this chunk
df_mosaich %>%
  select(DEMO1) %>%
  distinct() %>%
  mutate(number = as.numeric(as.factor(DEMO1))) %>%
  arrange(number)
```


### Question

1. Please interpret the results.
2. Which result relates to the research question _Is there any Swiss region, where women/men are especially happy?_ and how would you answer it based on the results?

### Answer

1. [YOUR ANSWER]
2. [YOUR ANSWER]


# TASK 2

## Prep

_Note_: In case you have renamed the folder/file-name (the names defined within the zip-file might be too long for Windows-systems), please adjust the path in the chunk below accordingly.

```{r}
# just run this chunk
df_act_raw <- readxl::read_xlsx('../data/a_workplace_acceptance_and_commitment_therapy_intervention_for_improving_healthcare_staff_psychological_distress/Data _ACT_for_NHS_RCT_Trial_OPS_Prudenzi_et_al_22.xlsx')
```

Take a look at data to see how it is structured.

```{r see_act_data}
# just run this chunk

df_act_raw %>%
  # limit to top 20 rows
  slice_head(n=20)
```

We transform the data into the so called [long](https://www.statology.org/long-vs-wide-data/)-format, because this is required by most analysis functions suitable for this task.

```{r}
# just run this chunk

df_act_long <- df_act_raw %>%
  # limit to required columns
  select(
    starts_with('Psychological_distress'),
    `Experimental_Control participant`,
    ID
  ) %>% 
  # wide to long
  pivot_longer(
    cols = starts_with('Psychological_distress'),
    names_to = 'time',
    values_to = 'psy_distress'
  ) %>%
  # clean strings
  mutate(
    time = stringr::str_replace(time, 'Psychological_distress_', ''),
    time = forcats::as_factor(time),
    # convert to factor and use new name
    group = forcats::as_factor(`Experimental_Control participant`)
  ) %>%
  # drop not needed column
  select(-`Experimental_Control participant`)
```


Take a look at data to see how it is structured now. 

Optional: you might want to check `levels(df_act_long$time)` as well to get the order of time levels. If it doesn't suit your needs you can use `forcats::fct_relevel()` to change it. Use `?forcats::fct_relevel()` to see its manual entry.

```{r see_act_long_data}
# just run this chunk

df_act_long %>%
  # limit to top 20 rows
  slice_head(n=20)
```


## Optional

Explore the data (this is optional for the homework).

```{r}
# ADD YOUR CODE


```


## Parametric



Conduct a parametric mixed ANOVA with `ez::ezANOVA()`. See its manual with `?ez::ezANOVA()`. You don't need to decompose any effects - just the omnibus test is correct.


```{r}
# FILL IN THE CORRECT VARIABLES

para_mixed <- ez::ezANOVA(
  df_act_long,
  wid = FILL_IN,
  dv = FILL_IN,
  within = FILL_IN,
  between = FILL_IN
)

para_mixed
```

### Question

1. Please interpret the results.
2. Do the results align with those in the paper? (you don't need to check the partial $\eta$² or the decomposed effects)

### Answer

1. [YOUR ANSWER]
2. [YOUR ANSWER]


## Robust

## Robust

Choose a robust version of the parametric test above in the [WRS2](https://cran.r-project.org/web/packages/WRS2/WRS2.pdf) and carry it out.


```{r robust_mixed}
# YOUR CODE


```


### Question

1. Please interpret the results.
2. Do the results align with those in the paper?


### Answer

1. [YOUR ANSWER]
2. [YOUR ANSWER]



# At last

* check that you have
  * added your name at the top of the document as author
  * added all missing code segments
  * answered all questions
* knit to PDF or HTML
* upload to moodle