---
title: "Mean differences - Practical examples"
author: "Bernhard Piskernik"
date: "2022/09/29"
output: 
  ioslides_presentation:
        css: ../style.css
        incremental: true
---



```{r setup, include=FALSE}
library(tidyverse)
library(plotly)
library(kableExtra)
library(sf)
options(warn=-1)
options("kableExtra.html.bsTable" = T)
theme_set(theme_minimal())
```


```{r helper, include=FALSE}
toTable <- function(df){
  df %>% kable() %>% kable_styling()
  }
```


```{r data_load, include=FALSE}
# retrieved shape files from https://www.bfs.admin.ch/asset/en/22484210
regions <- st_read('../data/ag-b-00.03-875-gg22/ggg_2022_LV95/shp/g1r22.shp', quiet=TRUE)

# retrieve MOSAiCH data from https://doi.org/10.48573/t659-e039
mosaich <- haven::read_sav('../data/MOSAiCH_2021/swissubase_2033_1_0/2033_MOSAiCH2021_Data_E_v1.0.0.sav') %>%
  # use labels instead of values for Nuts2
  mutate(Nuts2 = haven::as_factor(Nuts2))

df_1f <- mosaich %>%
  # reduce to needed variables
  select(Nuts2, H1) %>%
  # remove missings
  drop_na() %>%
  # remove missing factor levels
  mutate(Nuts2 = forcats::fct_drop(Nuts2))
```



## One factoral design {.build}

Research question: *Does happiness differ between the Swiss regions?*




```{r,  echo=FALSE, message=FALSE, fig.height=4}
p <- ggplot()+
  geom_sf(data = regions, aes(fill = GRNAME)) +
  annotate(
    geom = 'text',
    label = as.character(expression(paste("\U1F603", "?"))),
    parse = TRUE, 
    size = 25,
    x = 2670000,
    y = 1200000
  )

p
```


## Data {.build}

Source: [MOSAiCH 2021. Measurement and Observation of Social Attitudes in Switzerland. Study on Health and Health Care and related topics](https://doi.org/10.48573/t659-e039)

Variables:

* `Nuts2`: _Large Regions_ 
* `H1` (variable name not a hypothesis): _Q1 How happy or unhappy_ [1 Completely happy - 7 Completely unhappy]

`H1` is obviously ordinal - can mean even be appropriate?

## Hypotheses {.build .flexbox .vcenter}

Hypothesis 1: The respondents from the 7 regions reported different mean happiness levels.


Hypothesis 2: Respondents from _Espace Mittelland_ reported higher mean happiness levels than _Zentralschweiz_.


## Look at the data - numerical 

\renewcommand{\arraystretch}{2}
```{r,  echo=FALSE, message=FALSE}
df_1f %>%
  group_by(Nuts2) %>%
  summarise(
    n = n(),
    mean = mean(H1) %>% round(2),
    trimmed10 = mean(H1, trim=.10) %>% round(2),
    median = median(H1) %>% round(2),
    sd = sd(H1) %>% round(2),
    var = var(H1) %>% round(2),
    skew = moments::skewness(H1) %>% round(2),
    kurt = moments::kurtosis(H1) %>% round(2)
  ) %>%
  kable(booktabs = TRUE, linesep = "\\addlinespace") %>%
  kable_styling(font_size = 22, latex_options = "striped")
```

## Look at the data - graphical {.build}

```{r,  echo=FALSE, message=FALSE, fig.height=4}
p <- df_1f %>%
  ggplot(aes(x=Nuts2, y=H1, fill=Nuts2)) +
    geom_boxplot() +
    theme(legend.position = "none")  +
    scale_x_discrete(guide = guide_axis(angle=30)) 

ggplotly(p) %>%
  config(displayModeBar = FALSE)
    
```

**Box plots** are excellent to display distributions.<br>Why are they not a good choice in case?

## Look at the data - graphical {.build}

```{r,  echo=FALSE, message=FALSE, fig.height=4}
p <- df_1f %>%
  ggplot(aes(x=H1, y = ..density.., fill=Nuts2)) +
    geom_histogram() +
    facet_wrap(~Nuts2, nrow=2) +
    theme(legend.position = "none")  

ggplotly(p) %>%
  config(displayModeBar = FALSE)
    
```

**WARNING**: depending on the bin size **histograms** can be misleading. 

## Look at the data - graphical {.build}

```{r,  echo=FALSE, message=FALSE, fig.height=4}
p <- df_1f %>%
  ggplot(aes(sample=H1, color=Nuts2)) +
    stat_qq(distribution=qnorm) + 
    stat_qq_line(distribution=qnorm) +
    facet_wrap(~Nuts2, nrow=2) +
    theme(legend.position = "none")  


ggplotly(p) %>%
  config(displayModeBar = FALSE)
    
```


Quantile-Quantile-plots are a great way to compare the sample distribution to a theoretical distribution. Ideally, the points would match the line.

Why do we see a stair pattern?

## Look at the data - graphical | add some random noise (normal [0, 0.5]) {.build .smaller}

```{r,  echo=FALSE, message=FALSE, fig.height=4}
p <- df_1f %>%
  mutate(H1 = H1 + rnorm(length(H1),0,0.5)) %>%
  ggplot(aes(sample=H1, color=Nuts2)) +
    stat_qq(distribution=qnorm) + 
    stat_qq_line(distribution=qnorm) +
    facet_wrap(~Nuts2, nrow=2) +
    theme(legend.position = "none")  


ggplotly(p) %>%
  config(displayModeBar = FALSE)
    
```

## Analysis - parametric | Omnibus {.smaller .build  #reduceTopMarginCode}


```{r, strip.white = TRUE}
oneway.test(H1~Nuts2,var.equal=FALSE, data=df_1f)
```


[Levine and Hullett (2002)](https://doi.org/10.1111/j.1468-2958.2002.tb00828.x) recommend Ω² or η² as **effect size** for ANOVAs.

* partial η² (used by SPSS) strongly depends on the variability of the residuals
* η² biased e.g. when _n_ is small or there are many levels

```{r}
aov(H1~Nuts2, data=df_1f) %>% effectsize::omega_squared(verbose=F) %>% toTable()
```


Hypothesis 1: The respondents from the 7 regions reported different mean happiness levels. --> Null-Hypothesis can be rejected, but the effect is minimal

## Analysis - parametric | Contrasts {.smaller .build}

```{r}
f1_lm <- lm(H1~Nuts2, data=df_1f)
f1_emm <- emmeans::emmeans(f1_lm, 'Nuts2', data=df_1f)
emmeans::test(
  emmeans::contrast(
    f1_emm, 
    list(ac1=c(0, 1, 0, 0, 0, -1, 0)) # this list can contain multiple contrasts
    ),
  adjust='none'
  )
```



