---
title: "Mi&nbsp;sing D&nbsp;ta"
author: "please insert your names here"
date: "`r Sys.Date()`"
output: html_document
---

# Training task (not graded)

* Repeat today's analysis, but with increase additional missing rate from 10 to 25%
* What do you notice?

The next few code chunks load the required packages and the data, and create 25% missingness.

The rest is up to you. Try it yourself - if it doesn't work you can copy & paste from the session's rmd-file (no adjustments are needed, because only the missing percentage was increased). But if you use c&p - think about what each code chunk does!

```{r setup}
# just run this chunk
library(tidyverse)
library(magrittr)
library(plotly)
library(kableExtra)
options(warn=-1)
options("kableExtra.html.bsTable" = T)
theme_set(theme_minimal())
```


```{r data_load}
# just run this chunk

# NOTE: we reuse the data from session 3: multiple regression
# retrieve data of project "Materialism & personality" from https://doi.org/10.3886/E101900V1
# attention: labels are Polish, so some transformation is necessary to facilitate work

df_mat <- haven::read_sav('../data/materalism_and_personality/Study_III.sav') %>%
  # use labels instead of values for Gender & syt_rodz
  mutate(
    GENDER = haven::as_factor(GENDER),
    syt_rodz = haven::as_factor(syt_rodz),
    # recode to English labels
    syt_rodz = forcats::fct_recode(
      syt_rodz,
      'single'='osoba samotna',
      'civil partnership'='w związku partnerskim',
      'married'='w związku małżeńskim',
      'other'='inne'
      )
    ) %>%
  # rename columns 1: all to lower case
  rename_all(tolower) %>%
  # rename columns 2: use full, English words
  rename(
    marital_status = syt_rodz,
    materialism = mvs_total,
    neuroticism = neurot,
    life_satisfaction = swls
  ) %>%
  # limit to used columns
  select(
    gender, marital_status, materialism, neuroticism, life_satisfaction
  )
  
```




The data has already some missing, but we will add an extra 25% per variable.

```{r}
# just run this chunk

set.seed(42)
mr <- 0.25
df <- purrr::map_df(df_mat, function(x) 
  {x[sample(c(TRUE, NA), prob = c(1-mr, mr), size = length(x), replace = TRUE)]})
df %>% head()
```

# Identify

```{r}
# ADD YOUR CODE

```

```{r}
# ADD YOUR CODE

```

# Examine

```{r}
# ADD YOUR CODE

```

```{r}
# ADD YOUR CODE

```


# Deal with the missing data


```{r}
# ADD YOUR CODE

```

```{r}
# ADD YOUR CODE

```

# My observations if the missing-rate increases to 25%

[YOUR ANSWER]



# At last (optional - if you want me to take a look at it)

* check that you have
  * added your name at the top of the document as author
  * added all missing code segments
  * answered all questions
* knit to PDF or HTML
* upload to moodle