---
title: "Mi&nbsp;sing D&nbsp;ta"
author: "Be&nbsp;nhard Pisk&nbsp;&nbsp;&nbsp;&nbsp;k"
date: "2022-12-15"
output: 
  ioslides_presentation:
        css: ../style.css
        incremental: true
        self_contained: true
---


```{r setup, include=FALSE}
library(tidyverse)
library(magrittr)
library(plotly)
library(kableExtra)
options(warn=-1)
options("kableExtra.html.bsTable" = T)
theme_set(theme_minimal())
```

```{r helper, include=FALSE}
toTable <- function(df){
  df %>% kable() %>% kable_styling()
  }
```

```{r data_load, include=FALSE}
# NOTE: we reuse the data from session 3: multiple regression
# retrieve data of project "Materialism & personality" from https://doi.org/10.3886/E101900V1
# attention: labels are Polish, so some transformation is necessary to facilitate work

df_mat <- haven::read_sav('../data/materalism_and_personality/Study_III.sav') %>%
  # use labels instead of values for Gender & syt_rodz
  mutate(
    GENDER = haven::as_factor(GENDER),
    syt_rodz = haven::as_factor(syt_rodz),
    # recode to English labels
    syt_rodz = forcats::fct_recode(
      syt_rodz,
      'single'='osoba samotna',
      'civil partnership'='w związku partnerskim',
      'married'='w związku małżeńskim',
      'other'='inne'
      )
    ) %>%
  # rename columns 1: all to lower case
  rename_all(tolower) %>%
  # rename columns 2: use full, English words
  rename(
    marital_status = syt_rodz,
    materialism = mvs_total,
    neuroticism = neurot,
    life_satisfaction = swls
  ) %>%
  # limit to used columns
  select(
    gender, marital_status, materialism, neuroticism, life_satisfaction
  )
  
```



## The Basic Problem {.build}

* Missing data is usual and not an exception.
* Usually no description, analyis, or even acknowledgment of missing data.
* That is bad and you shouldn't do that 👨‍🏫.
* Conclusions may (and do) change when missing data is accounted for.

Consequences of missing data:

* lower power
* bigger standard errors and confidence intervals
* biased results


## Steps in dealing missing data {.build}

1. Identify the missing data.
2. Examine (the causes of) the missing data.
3. Deal with the missing data.

Unfortunately, (1) is the only unambiguous step.

For those really interested: Check out the standard work by [Little & Rubin (2002)](https://onlinelibrary.wiley.com/doi/book/10.1002/9781119013563).


## A classification of missing data {.build .smaller}

**Missing completely at random MCAR** 

* the presence of missing data on a variable is unrelated to any other observed or unobserved variable
* easy to handle, but rare

**Missing at random MAR**

* the presence of missing data on a variable is related to other observed variables but **not** to its **own** unobserved value

**Not missing at random NMAR**

* the presence of missing data is NOT systematic or predictable using the other information we have but also isn’t missing randomly (related to its own unobserved value)
* analysis of NMAR data is complex (and beyond the scope of this course)
* often the best way of dealing with this is to try to collect more information about why the data is missing


# Let's continue with an example

## Data (same as in session 3){.build .smaller}

Source: Górnik-Durose, Malgorzata E. Materialism & personality. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 2020-04-15. https://doi.org/10.3886/E101900V1

Publication: Górnik-Durose, M. E. (2020). Materialism and well-being revisited: The impact of personality. _Journal of Happiness Studies: An Interdisciplinary Forum on Subjective Well-Being_, _21_(1), 305–326. https://doi.org/10.1007/s10902-019-00089-8

The data has already some missings, but we will add an extra 10% per variable.

```{r}
set.seed(42)
mr <- 0.1
df <- purrr::map_df(df_mat, function(x) 
  {x[sample(c(TRUE, NA), prob = c(1-mr, mr), size = length(x), replace = TRUE)]})
df %>% head()
```

# 1. Identify the missing data.

## Tabular {.smaller}

```{r, message=FALSE, results='hide'}
dv = 'life_satisfaction'
iv = c('gender', 'marital_status', 'materialism', 'neuroticism')
df %>% finalfit::ff_glimpse(dv, iv)
```

```{r, echo=FALSE}
ffg <- df %>% finalfit::ff_glimpse(dv, iv)
ffg$Continuous %>% toTable()
```

```{r, echo=FALSE}
ffg$Categorical %>% toTable()
```

## Graphical

```{r}
df %>% finalfit::missing_plot() %>% ggplotly()
```

# 2. Examine (the causes of) the missing data.

## Things you want to answer

* What percentage of the data is missing?
* Are the missing data concentrated in a few variables or widely distributed?
* Do the missing values appear to be random?
* Does the covariation of missing data with each other or with observed data suggest a possible mechanism that’s producing the missing values?


## Visual Patterns {.smaller}

```{r}
df %>% VIM::aggr(prop=FALSE, numbers=TRUE, cex.axis=.7 )
```

## Visual Patterns

* Are any variables or variable combinations missing more often?
* Is there something special about them?
  - sensitive data
  - procedure (e.g. at the end of long questionnaire)
  - ...
  
  
## Associations between missing and observed data (graphical) {.smaller .reduceTopMarginCode}

```{r, message=FALSE, warning=FALSE}
df %>% finalfit::missing_pairs(dv, iv)
```

## Associations between missing and observed data (numerical) {.smaller}

```{r, eval=FALSE}
df %>% finalfit::missing_compare(dv, iv)
```


```{r, echo=FALSE}
df %>% 
  finalfit::missing_compare(dv, iv) %>% 
  toTable()
```

Take _p_-values with a grain of salt ($\alpha$-inflation, not necessarily adequate test).

## Omnibus test for testing MCAR {.build}

[Little's (1988)](https://doi.org/10.1080/01621459.1988.10478722) test statistic assesses if data is missing completely at random (MCAR). The null hypothesis in this test is that the data is MCAR, and the test statistic is a chi-squared value.

```{r}
naniar::mcar_test(df)
```

Rather rely on your understanding of the mechanism causing missings than on this test to decide whether your data is MCAR.


# 3. Deal with the missing data.

## Approaches

* rational
* complete-case analysis (listwise deletion)
* multiple imputation
* other approaches
  - FIML
  - pairwise deletion
  - single imputation
  - simple (nonstochastic) imputation
  - ...
  
  
## Rational approach

* use mathematical or logical relationships among variables to attempt to fill in or recover missing value
* e.g.: gender from first name, age from birth year, income from job, ...
* typically requires creativity and thoughtfulness
* data recovery may be exact or approximate


## Complete-case analysis (listwise deletion)

* default in most software packages (most often used)
* if MCAR unbiased, but smaller sample size with all consequences (reduces statistical power)
* if not MCAR biased/skewed results


## Multiple imputation

* 3 steps: 
  - **impute**: distributions of the missing values are estimated (via regression techniques) and drawing from them multiple complete data sets (10-20 are typical) are created
  - **analyze**: analysis is done with each data set
  - **pool**: results from all data sets are pooled
* frequently the method of choice for complex missing-values problems

## Multiple imputation - Tips

* keep auxiliary variables (neither DV nor IV) in data set (might help predict the missing values) 
* don't be fancy: impute your DV and also keep it in the analysis (see [Kontopantelis et al., 2017](https://doi.org/10.1186/s12874-016-0281-5))
* double check the scale of your variables and use adequate regression technique (are categorical defined as categorical?)

## Multiple imputation - Impute {.build .smaller}



```{r, results='hide'}
imp <- mice::mice(df, m = 5, seed = 42) # use m=20
```

```{r}
summary(imp)
```


## Multiple imputation - Analyze {.build .smaller}

```{r}
fit_mi <- with(imp, lm(life_satisfaction ~ neuroticism + materialism + marital_status))
```

```{r, echo=FALSE}
map(fit_mi$analyses, function(x) x$coefficients) %>% bind_rows() %>% toTable()
```


## Multiple imputation - Pool {.build .smaller}

```{r}
fit_mi %>% mice::pool() %>% summary()
```

## Combine bootstrapping with MI {.build .smaller .reduceTopMarginText}

create bootstrapped & MI data sets - Number of data sets = nBoot*nImp

```{r, eval=FALSE, class.source='bottomMargin-5'}
boot_imp <- bootImpute::bootMice(df, nBoot=100, nImp=5, seed=42) # use >= 1000 & 20
```

```{r, cache=TRUE, results='hide', echo=FALSE, message=FALSE, warning=FALSE}
boot_imp <- bootImpute::bootMice(df, nBoot=100, nImp=5, nCores=4, seed=42, printFlag=FALSE) # use >= 1000 & 20
```

analyze and pool

```{r, class.source='bottomMargin-5'}
# wrapper to analyze a data set
get_coefficients <- function(df) {
  coef(lm(life_satisfaction ~ neuroticism + materialism + marital_status, df))}
fit_boot_mi <- bootImpute::bootImputeAnalyse(boot_imp, get_coefficients)
```

For more information see e.g. [Bartlett and Hughes (2020)](https://doi.org/10.1177/0962280220932189).

## Other approaches {.build .smaller .reduceTopMarginText}

FIML (full information maximum likelihood)

- equivalent results to MI (see e.g. [Li & Shi, 2021](https://doi.org/10.1037/met0000381))
- only within structural equation modeling

Pairwise deletion

  - different analyses use different samples (what's the sample size?)
  - biased results if not MCAR & indefinite correlation matrix
  
Single imputation

  - = MI with m=1 (no pooling)
  - underestimates Standard Errors

Simple (nonstochastic) imputation

  - one value for all
  - e.g. mean or median for continuous variables
  - own missing class for categorical variables
  - biased results if not MCAR
  - underestimates SEs (too little variance in the sample)
  
## Training task (not graded) {.flexbox .vcenter}

* Repeat today's analysis, but with increase additional missing rate from 10 to 25%
* What do you notice?

## Thank you for your attention! {.flexbox .vcenter}