---
title: "Multiple Regression - Homework"
author: "please insert your names here"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tasks

due till 2022-11-10

* use the data, research question and variables from the lecture on _multiple regression_
* build a _robust_ model (without interaction) using the function `MASS::rlm()`
* describe the differences you see compared to _model4_

# Preparations

## Libraries

Load required libraries

```{r load_libraries}
# just run this chunk

library(tidyverse)
library(MASS)
library(rsample)
library(sfsmisc)
```


## Data

1. Create an account at [openicpsr](https://www.openicpsr.org/) to get access to the data.
2. Make sure that the [Materialism & personality](https://doi.org/10.3886/E101900V1) data-set is extracted at _[your individual folder path]/2022_SA_advanced_statistical_methods/data_.


# Task

## Prep

Get data into shape.

```{r prepare_mosaich_data}
# just run this chunk

# retrieve data of project "Materialism & personality" from https://doi.org/10.3886/E101900V1
# attention: labels are Polish, so some transformation is necessary to facilitate work

df_mat <- haven::read_sav('../data/materalism_and_personality/Study_III.sav') %>%
  # use labels instead of values for Gender & syt_rodz
  mutate(
    GENDER = haven::as_factor(GENDER),
    syt_rodz = haven::as_factor(syt_rodz),
    # recode to English labels
    syt_rodz = forcats::fct_recode(
      syt_rodz,
      'single'='osoba samotna',
      'civil partnership'='w związku partnerskim',
      'married'='w związku małżeńskim',
      'other'='inne'
      )
    ) %>%
  # rename columns 1: all to lower case
  rename_all(tolower) %>%
  # rename columns 2: use full, English words
  rename(
    marital_status = syt_rodz,
    materialism = mvs_total,
    neuroticism = neurot,
    life_satisfaction = swls
  ) %>%
  # limit to used columns
  dplyr::select(
    gender, marital_status, materialism, neuroticism, life_satisfaction
  )
```

Take a look at data to see how it is structured

```{r see_mosaich_data}
# just run this chunk

df_mat %>%
  # limit to top 20 rows
  slice_head(n=20)
```

## Optional

Explore the data (this is optional for the homework). Tip you can use the code from the lecture as a starting point.

```{r}
# ADD YOUR CODE


```

## Model 4

Model 4 used standardized variables, so we need to standardize all numeric variables.


```{r standardize}
# just run this chunk
df_z <- df_mat %>% mutate_if(is.numeric, scale)
```

Now, we can create _Model 4_

```{r}
# just run this chunk
model4 <- lm(life_satisfaction ~ neuroticism + materialism + marital_status, data = df_z)

summary(model4)
```

## Robust with weighted least squares

Now it is your turn to build a _robust_ model using the function `MASS::rlm()`

```{r rlm}
# ADD YOUR CODE
set.seed(42)
model4_bt<- df_z %>% 
  rsample::bootstraps(1000) %>% 
  pull(splits) %>% 
  map_dfr(~rlm(life_satisfaction ~ neuroticism + materialism + marital_status, data = .)%>% 
            broom::tidy())
model4_bt %>% 
  group_by(term) %>% 
  dplyr::summarize(est_mean = mean(estimate),
    est_median = median(estimate),
    low=quantile(estimate, .025),
    high=quantile(estimate, .975))
```


_Note_: `rlm()` objects do not include p-values. You can get those, e.g., by bootstrapping or with the function `sfsmisc::f.robftest()` 


### Question

1. describe the differences you see compared to _model4_er?


### Answer

1. There are hardly any. The parameters are very similar and the same parameter are significant/not significant. This indicates that the OLS assumptions are met.



# At last

* check that you have
  * added your name at the top of the document as author
  * added all missing code segments
  * answered all questions
* knit to PDF or HTML
* upload to moodle